Wait-free Queues with Polygarithmic Step Complexity

### 2022-04-01 article/psuedocode.tex
fix fallback, add comments

\begin{itemize}
  \item \tt{\sl{int} num\sub{enq-left}, sum\sub{enq-left}}
  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \tt{num\sub{enq-left}}}
  \item \tt{\sl{int} num\sub{deq-left}, sum\sub{deq-left}}
  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \tt{num\sub{deq-left}}}
  \item \tt{\sl{int} num\sub{enq-right}, \tt{sum\sub{enq-right}}}
  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \tt{num\sub{enq-right}}}
  \item \tt{\sl{int} num\sub{deq-right}, \tt{sum\sub{deq-right}}}
  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \tt{num\sub{deq-right}}}
  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in the block}
  \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
  \item \tt{\sl{int} num, sum}
  \textsf{: total \# operations in block, prefix sum of \tt{num}}
  \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~index of the last subblock in the left and right child}
  \item \tt{\sl{int} group}
  \textsf{: id of the group of blocks including this propagated together, more precisely the value read from the node \tt{n}'s counter when propagating this block to the node \tt{n}.}
  \item \tt{\sl{int} order}
  \textsf{: the index of the block in the node containing it}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Root Block} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{int} size}
  \textsf{: size of queue after this block's operations finish}
  \item \tt{\sl{int} sum\sub{non-null deq}}
  \textsf{: count of non-null dequeus up to this block}
  \item \tt{\sl{int} num\sub{done}}
  \textsf{: number of finished operations in the block}
\end{itemize}

\Function{void}{Enqueue}{\sl{Object} e}
\State \tt{block b= \Call{new}{\sl{block}}}
\State \tt{b.element= e}
\State \tt{b.num\sub{enq}=1}
\State \tt{b.sum\sub{enq}= this.blocks[leaf.head].sum\sub{enq}+1}
\State \Call{Append}{b}
\EndFunction{Enqueue}

\Function{Object}{Dequeue()}{}
\State \tt{block b= \Call{new}{\sl{block}}}
\State \tt{b.element= null}
\State \tt{b.num\sub{deq}=1}
\State \tt{b.sum\sub{deq}= this.blocks[leaf.head].sum\sub{deq}+1}
\State \Call{Append}{b}
\State \tt{<i, b\sub{i}>=} \Call{Index}{this.leaf, this.leaf.head, 1} \cmt{\tt{i} is the order in the root among all dequeues, of the dequeue in the last block in the process's leaf. \tt{b\sub{i}} is the block in the root containing it.}
\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1.} \label{deqRest}
\If{\tt{index\sub{response}!=-1}}
\State \tt{output= null}
\State \tt{b\sub{i}.num\sub{done}= b\sub{i}.num\sub{done}+1}
\If{\tt{b\sub{r}.num\sub{done}==b\sub{r}.num}}\cmt{become old}
\State \tt{this.leaf.last\sub{done}= b\sub{r}}
\EndIf
\Else
\State \tt{output= \Call{Get}{res}}
\State \tt{b\sub{r}= root.blocks.get(enq, index\sub{response})}\cmt{block in the root contains response enqueue.}
\State \tt{b\sub{i}.num\sub{done}= b\sub{i}.num\sub{done}+1}
\State \tt{b\sub{r}.num\sub{done}= b\sub{r}.num\sub{done}+1}
\If{\tt{b\sub{r}.num\sub{done}==b\sub{r}.num}}\cmt{become old}
\State \tt{this.leaf.last\sub{done}= b\sub{r}}
\ElsIf {\tt{b\sub{i}.num\sub{done}==b\sub{i}.num}}
\State \tt{this.leaf.last\sub{done}= b\sub{i}}
\EndIf
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{int}{ComputeDeqRes}{int i, int b}\cmt{Computes head of the queue when \tt{i}th dequeue in \tt{b}th block occurs. The dequeue should return the argument of the head enqueue.}
\If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{-1}
\Else{ \Return \tt{root.blocks[b-1].sum\sub{non-null deq} + i}}
\EndIf
\EndFunction{ComputeDeqRes}

@@

\Function{void}{Append}{\sl{block} b} 
\State \tt{b.group= this.leaf.head}
\State \tt{l\sub{pid}.blocks[this.leaf.head]= b}
\State \tt{this.leaf.head+=1}
\State \Call{Propagate}{this.leaf.parent} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\sl{node} n}
\If{\tt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\If{\tt{n.parent} \bf{is not} \tt{null}} 
\State \Call{Propagate}{n.parent}
\EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\sl{node} n}
\State \tt{h= n.head}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{c= n.counter}
\State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}}
\State \tt{new.group= c}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{(\tt{n} \bf{is} \tt{root} \bf{and} \tt{root.blocks.append(new)}) \bf{or} \\ (\tt{n} \bf{is not} \tt{root} \bf{and} \Call{CAS}{n.blocks[h], null, new})} \cmt{\it{space in he first of the new line?}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{n.dir.super[c\sub{dir}], null, h+1}}
\State \tt{\Call{CAS}{n.dir.counter, c\sub{dir}, c\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{n.head, h, h+1}}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{n.head, h, h+1}}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Function{element}{Get}{\sl{int} i}
\cmt{Returns $i$th Enqueue.}
\If{\tt{i }\bf{is }\tt{null}}
\State \Return \tt{null}
\EndIf
\State \tt{res= root.blocks.get(enq, i).order}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
\EndFunction{Get}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{node} n, \sl{int} i} 
\Statex\cmt{Creates a block to insert into \tt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
\State \tt{block b= \Call{new}{\sl{block}}}
\If{\tt{n} \bf{is} \tt{root}}
\State \tt{block b= \Call{new}{\sl{root block}}}
\EndIf
\State \tt{b.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{lastIndex= n.dir.head} \label{lastLine}
\State \tt{prevIndex= n.blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{lastBlock= n.dir.blocks[lastIndex]}
\State \tt{prevBlock= n.dir.blocks[prevIndex]}
\State \tt{c\sub{dir}= n.dir.counter}
\State \tt{b.end\sub{dir}= lastIndex}
\State \tt{b.num\sub{enq-dir}= lastBlock.sum\sub{enq} - prevBlock.sum\sub{enq}}
\State \tt{b.num\sub{deq-dir}= lastBlock.sum\sub{deq} - prevBlock.sum\sub{deq}}
\State \tt{b.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
\State \tt{b.sum\sub{deq-dir}= n.blocks[i-1].sum\sub{deq-dir} + b.num\sub{deq-dir}}
\EndFor
\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{n.parent} \bf{is} \tt{null}}
\State \tt{b.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].size - b.num\sub{enq}, 0)}
\EndIf

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of node \tt{n} has propagated up to the root and \tt{i}th dequeue resides in node \tt{n} is in block \tt{b} of node \tt{n}.}
\Function{<int, int>}{Index}{\sl{node} n, \sl{int} b, \sl{int} i} \cmt{Returns the order in the root among dequeus, of ith dequeue in bth block of node n.}
\If{\tt{n {\keywordfont is} root}} \Return \tt{root.blocks.get(order==b-1).sum\sub{deq}+i, b}
\Else
\State \tt{dir= (n.parent.left==n)? left: right}
\State \tt{superBlock= \Call{BSearch}{n.parent, n.sum\sub{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
\If{\tt{dir {\keywordfont is} left}}
\State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq-right}}
\Else \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq} + n.blocks[superBlock].sum\sub{deq-left}}
\EndIf
\State \Return\Call{Index}{n.parent, superBlock, i}
\EndIf
\EndFunction{Index}

@@

\Statex $\blacktriangleright$ \tt{\sl{PRBTree[rootBlock]}}
  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}. \tt{append(b, key)} returns \tt{true} in case successful.

\Function{void}{RBTAppend}{block b} \cmt{\textsf{adds block b to the \tt{root.blocks}}}
\State \tt{step= root.head}
\If{\tt{step\%$p^2$==0}}
\State \tt{Help()}
\State \tt{CollectGarbage()}
\EndIf
\State \tt{b.age= 0}
\State \Return \tt{root.blocks.append(b, b.order)}
\EndFunction{RBTAppend}
\Statex

\Function{void}{Help}{}\cmt{Helps pending operations}
\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{how to iterate over them?}}
\State{\tt{last= l.head-1}}
\If{\tt{l.blocks[last]} \bf{is not} \tt{null}}
\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
\State \tt{goto \ref{deqRest} with these values <>} \cmt{run \tt{Dequeue()}  for \tt{l.ops[last]} after Propagate(). \it{TODO}}
\State \tt{l.responses[last]= response}
\EndIf
\EndIf
\EndFor
\EndFunction{Help}
\Statex

\Function{void}{CollectGarbage}{}\cmt{Collects the old root blocks.}
\State \tt{l=FindYoungestOld(Root.Blocks.root)}
\State \tt{t1,t2= RBT.split(l)}
\State \tt{RBTRoot.CAS(t2.root)}
\EndFunction{CollectGarbage}

@@

\Function{Block}{FindYoungestOld}{b}
\For{\tt{leaf l} \bf{in leaves}}
\State\tt{max= Max(l.maxOld, max)}
\EndFor
\State\Return\tt{max} \cmt{This snapshot suffies.}
\EndFunction{findYoungestOld}

@@

\If{a dequeue cannot find the root block}

>>>>

\begin{itemize}
  \item \tt{\sl{int} num\sub{enq-left}, sum\sub{enq-left}}
  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \tt{num\sub{enq-left}}}
  \item \tt{\sl{int} num\sub{deq-left}, sum\sub{deq-left}}
  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \tt{num\sub{deq-left}}}
  \item \tt{\sl{int} num\sub{enq-right}, \tt{sum\sub{enq-right}}}
  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \tt{num\sub{enq-right}}}
  \item \tt{\sl{int} num\sub{deq-right}, \tt{sum\sub{deq-right}}}
  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \tt{num\sub{deq-right}}}
  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in the block}
  \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
  \item \tt{\sl{int} num, sum}
  \textsf{: total \# operations in block, prefix sum of \tt{num}}
  \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~index of the last subblock in the left and right child}
  \item \tt{\sl{int} group}
  \textsf{: id of the group of blocks including this propagated together, more precisely the value read from the node \tt{n}'s counter when propagating this block to the node \tt{n}.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Root Block} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{int} size}
  \textsf{: size of queue after this block's operations finish}
  \item \tt{\sl{int} sum\sub{non-null deq}}
  \textsf{: count of non-null dequeus up to this block}
  \item \tt{\sl{int} num\sub{finished}}
  \textsf{: number of finished operations in the block}
    \item \tt{\sl{int} order}
  \textsf{: the index of the block in the node containing it. Useful in the root since in the PBRT we do not keep indices in another way.}
\end{itemize}

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
\State \tt{block b= \Call{new}{\sl{leaf block}}}
\State \tt{b.element= e}
\State \tt{b.num\sub{enq}=1}
\State \tt{b.sum\sub{enq}= this.leaf.blocks[this.leaf.head].sum\sub{enq}+1}
\State \Call{Append}{b}
\EndFunction{Enqueue}

\Statex
\Function{Object}{Dequeue()}{}
\State \tt{block b= \Call{new}{\sl{leaf block}}} \cmt{Creates a null element block, appends it to the tree, computes its order among operations, then computes its response index if it exists and returns the response's element.}
\State \tt{b.element= null}
\State \tt{b.num\sub{deq}=1}
\State \tt{b.sum\sub{deq}= this.leaf.blocks[this.leaf.head].sum\sub{deq}+1} \cmt{\tt{this} is the current running process}
\State \Call{Append}{b}
\State \tt{<i, b\sub{i}>=} \Call{Index}{this.leaf, this.leaf.head, 1} \cmt{\tt{i} is the order in the root among all dequeues, of the dequeue in the last block in the process's leaf. \tt{b\sub{i}} is the index of the block in the root containing it. Since only one invocation of \tt{dequeue} is running by \tt{this} process at one time we are allowed to used \tt{this.leaf.head} safely.}
\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
\If{\tt{index\sub{response}!=-1}}
\State \tt{output= null}
\State \tt{b\sub{deq}=root.blocks[b\sub{i}]}
\State \tt{b\sub{deq}.num\sub{finished}.inc()} \cmt{shared counter}
\If{\tt{b\sub{deq}.num\sub{finished}==b\sub{deq}.num}}\cmt{all the operations in the block containing the dequeue are finished.}
\State \tt{this.leaf.last\sub{done}= b\sub{i}}
\EndIf
\Else
\State \tt{output= \Call{Get}{index\sub{response}}}
\State \tt{b\sub{r}= root.blocks.get(enq, index\sub{response}).order}\cmt{index of the block in the root contains response enqueue.}
\State \tt{b\sub{enq}=root.blocks[b\sub{r}]}
\State \tt{b\sub{enq}.num\sub{finished}.inc()}
\State \tt{b\sub{deq}.num\sub{finished}.inc()}
\If{\tt{b\sub{enq}.num\sub{finished}==b\sub{enq}.num}}\cmt{become done}
\State \tt{this.leaf.last\sub{done}= b\sub{r}}
\ElsIf {\tt{b\sub{deq}.num\sub{finished}==b\sub{deq}.num}} \cmt{\tt{b\sub{deq}} comes after \tt{b\sub{enq}}.}
\State \tt{this.leaf.last\sub{done}= b\sub{i}} \cmt{\tt{this.leaf.last\sub{done}} is an increasing value.}
\EndIf
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\sl{block} b} 
\State \tt{b.group= this.leaf.head} \label{appendStart} \cmt{Only this block is propagated from the leaf by itself.}
\State \tt{this.leaf.blocks[this.leaf.head]= b} 
\State \tt{this.leaf.head+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
\State \Call{Propagate}{this.leaf.parent} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\sl{node} n}
\If{\tt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n} \cmt{Lemma Double Refresh}
\EndIf
\If{\tt{n} \bf{is not} \tt{root}} \cmt{To check anode is the root we can check its index if the tree is implemented by an array or check if \tt{n.parent} is not \tt{null}.} 
\State \Call{Propagate}{n.parent}
\EndIf
\EndFunction{Propagate}

@@

\Function{element}{Get}{\sl{int} i}
\cmt{Returns $i$th Enqueue.}
\State \tt{res= root.blocks.get(enq, i).order}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
\EndFunction{Get}

@@

\Function{int}{ComputeDeqRes}{int i, int b}\cmt{Computes the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns the index of the the head of the queue or -1 if queue is empty.}
\If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{-1}
\Else{ \Return \tt{root.blocks[b-1].sum\sub{non-null deq} + i}}
\EndIf
\EndFunction{ComputeDeqRes}

\Function{boolean}{Refresh}{\sl{node} n}
\State \tt{h= n.head}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{c= n.counter}
\State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}} \cmt{\tt{c\sub{left}, c\sub{right}} are the values read from \tt{n}'s children's counters.}
\State \tt{new.group= c}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{(\tt{n} \bf{is} \tt{root} \bf{and} \tt{root.blocks.append(new)}) \bf{or} \\ (\tt{n} \bf{is not} \tt{root} \bf{and} \Call{CAS}{n.blocks[h], null, new})} \cmt{\it{how to put space in he first of the new line?}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{n.dir.super[c\sub{dir}], null, h+1}} \cmt{Superblock's Lemma}
\State \tt{\Call{CAS}{n.dir.counter, c\sub{dir}, c\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{n.head, h, h+1}}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{n.head, h, h+1}} \cmt{Even if another process wins, help to increase the head. It might fell sleep before increasing.}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{node} n, \sl{int} i} 
\Statex\cmt{Creates a block to insert into \tt{n.blocks[i]}. Returns the created block as well as values read from each child counter field. The values are used for incrementing children's counters if the block was appended to \tt{n.blocks} successfully. \it{Does it need help? I think no but in that case we do not have to pass these values to the calling line.}}
\State \tt{block b= \Call{new}{\sl{block}}}
\If{\tt{n} \bf{is} \tt{root}}
\State \tt{b= \Call{new}{\sl{root block}}}
\EndIf
\State \tt{b.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= n.dir.head} \label{lastLine}
\State \tt{index\sub{prev}= n.blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{block\sub{last}= n.dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= n.dir.blocks[index\sub{prev}]} \cmt{\tt{n.dir.blocks[index\sub{prev}..index\sub{last}]} are merged to one block.}
\State \tt{c\sub{dir}= n.dir.counter}
\State \tt{b.end\sub{dir}= index\sub{last}}
\State \tt{b.num\sub{enq-dir}= block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{b.num\sub{deq-dir}= block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\State \tt{b.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
\State \tt{b.sum\sub{deq-dir}= n.blocks[i-1].sum\sub{deq-dir} + b.num\sub{deq-dir}}
\EndFor
\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{n} \bf{is} \tt{root}}
\State \tt{b.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].size - b.num\sub{enq}, 0)}
\EndIf

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of node \tt{n} has propagated up to the root and \tt{i}th dequeue resides in node \tt{n} is in block \tt{b} of node \tt{n}.}
\Function{<int, int>}{Index}{\sl{node} n, \sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
\If{\tt{n {\keywordfont is} root}} \Return \tt{i, b}
\Else
\State \tt{dir= (n.parent.left==n)? left: right} \cmt{check \tt{n} is a left or a right child}
\State \tt{superBlock= \Call{BSearch}{n.parent, n.sum\sub{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}} \cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} left}}
\State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq-right}}
\Else \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq} + n.blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from \tt{n}'s right child}
\EndIf
\State \Return\Call{Index}{n.parent, superBlock, i}
\EndIf
\EndFunction{Index}

@@

\Statex $\blacktriangleright$ \tt{\sl{PRBTree[rootBlock]}}
  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
  \tt{append(b, key)} returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.

\Function{void}{RBTAppend}{block b} \cmt{\textsf{adds block b to the \tt{root.blocks}}}
\State \tt{step= root.head}
\If{\tt{step\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root. Used in lemma's using the size of the PBRT.}
\State \tt{Help()}
\State \tt{CollectGarbage()}
\EndIf
\State \tt{b.num\sub{finished}= 0}
\State \Return \tt{root.blocks.append(b, b.order)}
\EndFunction{RBTAppend}
\Statex

\Function{void}{Help}{}\cmt{Helps pending operations}
\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{if the tree is implemented with an array we can iterate over the second half of the array.}}
\State{\tt{last= l.head-1}}
\cmt{\tt{l.blocks[last]} can not be \tt{null} because of lines \ref{appendStart}-\ref{appendEnd}.}
\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
\State \tt{goto \ref{deqRest} with these values <>} \cmt{run \tt{Dequeue()}  for \tt{l.ops[last]} after Propagate(). \it{TODO}}
\State \tt{l.responses[last]= response}
\EndIf
\EndFor
\EndFunction{Help}
\Statex

\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
\State \tt{t1,t2= RBT.split(order, s)}
\State \tt{RBTRoot.CAS(t2.root)}
\EndFunction{CollectGarbage}

@@

\Function{Block}{FindMostRecentDone}{b}
\For{\tt{leaf l} \bf{in leaves}}
\State\tt{max= Max(l.maxOld, max)}
\EndFor
\State\Return\tt{max} \cmt{This snapshot suffies.}
\EndFunction{findYoungestOld}

@@

\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
